### 2022-11-08 submission/manuscript.Rmd
First full draft of Materials and Methods

@@

***Without rarefaction, metrics of beta diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring beta diversity to sampling effort, I used the same null models used for studying the sensitivity of alpha diversity. Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample (e.g., Figure S3). First, I calculated the Jaccard distance coefficient between all pairs of communities within a dataset. The Jaccard distance coefficient is the fraction of OTUs that are unique to either community and does not account for the abundance of the OTUs. Jaccard distances were calcualted using the uncorrected OTU counts, with rarefaction, relative abundances, and following normalization using cumulative sum scaling (CSS) and SRS. Only the rarefied distances showed a lack of sensitivity to sampling effort (Figure 1B). Second, I analyzed the sensitivity of the Bray-Curtis distance coefficient, which is a popular metric that incorporates the abundance of each OTU. Similar to what I observed with the Jaccard coefficient, only the rarefied data showed a lack of sensitivity to sampling effort (Figure 1B). Third, I calcualted the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring OTUs in samples with zero counts (Robust CLR), adding a pseudocount of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudocount of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR). The Aitchison distances were all strongly sensitive to sampling effort (Figure 1B). Finally, I used the variance stabilization technique (VST) from DeSeq2 prior to calculating Euclidean distances. Again, there was a strong sensitivity to sampling effort (Figure 1B). Although Euclidean distances are not typically used on raw or rarefied count data in ecology, rarefied Euclidean distances were not sensitive to sampling effort. Across each of the beta diversity metrics and approaches used to account for uneven sampling effort and sparsity, rarefaction was the least sensitive approach to differences in sampling effort.

@@

**Richness-adjusted community model.** In the richness-adjusted community model, communities were randomly assigned to one of two simulated treatment groups. Communities in the first treatment group were generated by calculating the relative abundance of each OTU across all samples and using those values as the probability of sampling each OTU. This probability distribution was sampled until each sample had the same number of sequences that it did in the observed data. Samples in the second treatment group were generated by removing 3% of the OTUs from the dataset and recalculating the relative abundance of the remaining OTUs. This percentage was determined after empirically searching for a value that resulted in a large fraction of the randomizations yieleding a significant result across most of the studies. Sequences were sampled from the richness-adjusted community distrubtion until each sample had the same number of sequences that it did in the observed data. Under the richness-adjusted community model each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups. Because the construction of the richness-adjusted community model was a stochastic process, 100 replicates were geneated for each dataset.

@@

The inverse Simpson diversity was calculated as $1/D_{simpson}$. In both formulae, $n_i$ is the number of sequences in OTU $i$ and $N$ is the number of sequences in the sample. Rarefaction of richness, Shannon diversity and Inverse Simpson diversity values were carried out in mothur. Briefly, mothur calculates each value on a random draw of the same number of sequences from each sample and obtains a mean value based on 1,000 random draws. Scaled ranked subsampling (SRS) was used to normalize OTU counts to the size of the smallest sample in each dataset using the SRS R package (v.`r package_version("SRS")`)[@Beule2020]. Normalized OTU counts were used to calculate sample richness and Shannon and inverse Simpson diversity values using mothur. The non-parametric bias-corrected Chao1 and ACE richness estimators [@Chao2016] and a non-parametric estimator of the Shannon diversity [@Chao2003] were calculated using raw OTU counts with mothur. Parametric estimates of sample richness were calculated using the breakaway (BA) R package (v.`r package_version("breakaway")`)[@Willis2015]. The current analysis reports both the results from running default model selection procedure and the Poisson model. The default model selection returned either the Kemp, Negative Binomial, or Poisson models.

@@

These metrics were calculated using the relative abundance of each OTU using the vegdist function from vegan. The relative abundance was calculated as the number of sequences in the OTU (e.g., $n_{A,i}$) divided by the total number of sequences in the sample (e.g., $N_A$). Rarefied beta-diversity values were calculated using the avgdist function in vegan. Briefly, vegan's avgdist function calculates each pairwise dissimilarity index after obtaining a random draw of the same number of sequences from each sample. After obtaining 100 random draws it returns the mean value.

SRS normalized - Jaccard/BC
CSS normalized - Jaccard/BC
DeSeq2
Center-log ratio

**Power analysis.**
 The fraction of tests that yielded a significant test was a measure of the statistical power for the test.

**Test of statistical significance.**
Wilcox, PERMANOVA

**Reproducible data analysis.** 
mothur - 1.47.0

***doi: 10.1038/nmeth.2658.*** Should CSS be used in alpha diversity analysis? No. The richness and shannon diversity values were no different from those obtained with the raw abundances. From the paper, "The relative proportion of the features is unaffected by the normalization". CSS paper refers to relative abundance approach as total-sum normalization (TSN)

Robust CLR - remove zero counts and calculate CLR
One CLR - add a pseudocount of 1 to all observations
Nudge CLR - add a pseudocount of 1/total number of sequences in sample
Zero CLR - Impute the value of zeroes using zCompositions package

***https://edepot.wur.nl/547087.*** "Log- ratio PCA is designed to give results that are library size- independent. However, as we demonstrated mathematically and with examples based on simulated and real data, log- ratio PCA becomes library size- dependent, if there are many infrequent taxa (many zeroes) and library sizes differ largely. In this situation, the row centring used in log- ratio PCA brings an effect of r (the row mean of the log- transformed counts) in the clr- transformed matrix. Note that this effect is irrespective of whether or not these infrequent taxa are genuine or due to sequencing noise or allocation error. This library size dependence is unexpected in the sense that, after applying the clr, the transformed matrix is free of the effect of the row totals for strictly positive data (yij>0 for all i and j). We additionally demonstrate that library size variability causes a loss in power in detecting an effect of x with log- ratio RDA. If there is additionally a correlation between treatment and the library size, the type 1 error for detecting the effect of x can be seriously inflated."

https://academic.oup.com/bioinformatics/article/34/16/2870/4956011
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/
\vspace{10mm}

@@

@@

>>>>

***https://edepot.wur.nl/547087.*** "Log- ratio PCA is designed to give results that are library size- independent. However, as we demonstrated mathematically and with examples based on simulated and real data, log- ratio PCA becomes library size- dependent, if there are many infrequent taxa (many zeroes) and library sizes differ largely. In this situation, the row centring used in log- ratio PCA brings an effect of r (the row mean of the log- transformed counts) in the clr- transformed matrix. Note that this effect is irrespective of whether or not these infrequent taxa are genuine or due to sequencing noise or allocation error. This library size dependence is unexpected in the sense that, after applying the clr, the transformed matrix is free of the effect of the row totals for strictly positive data (yij>0 for all i and j). We additionally demonstrate that library size variability causes a loss in power in dextecting an effect of x with log- ratio RDA. If there is additionally a correlation between treatment and the library size, the type 1 error for detecting the effect of x can be seriously inflated."

https://academic.oup.com/bioinformatics/article/34/16/2870/4956011
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/

@@

***Without rarefaction, metrics of beta diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring beta diversity to sampling effort, I used the same null models used for studying the sensitivity of alpha diversity. Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample (e.g., Figure S3). First, I calculated the Jaccard distance coefficient between all pairs of communities within a dataset. The Jaccard distance coefficient is the fraction of OTUs that are unique to either community and does not account for the abundance of the OTUs. Jaccard distances were calcualted using the uncorrected OTU counts, with rarefaction, relative abundances, and following normalization using cumulative sum scaling (CSS) and SRS. Only the rarefied distances showed a lack of sensitivity to sampling effort (Figure 1B). Second, I analyzed the sensitivity of the Bray-Curtis distance coefficient, which is a popular metric that incorporates the abundance of each OTU. Similar to what I observed with the Jaccard coefficient, only the rarefied data showed a lack of sensitivity to sampling effort (Figure 1B). Third, I calcualted the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring OTUs in samples with zero counts (Robust CLR), adding a pseudocount of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudocount of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR). The Aitchison distances were all strongly sensitive to sampling effort (Figure 1B). Finally, I used the variance-stabilization transformation (VST) from DeSeq2 prior to calculating Euclidean distances. Again, there was a strong sensitivity to sampling effort (Figure 1B). Although Euclidean distances are not typically used on raw or rarefied count data in ecology, rarefied Euclidean distances were not sensitive to sampling effort. Across each of the beta diversity metrics and approaches used to account for uneven sampling effort and sparsity, rarefaction was the least sensitive approach to differences in sampling effort.

@@

**Richness-adjusted community model.** In the richness-adjusted community model, communities were randomly assigned to one of two simulated treatment groups. Communities in the first treatment group were generated by calculating the relative abundance of each OTU across all samples and using those values as the probability of sampling each OTU. This probability distribution was sampled until each sample had the same number of sequences that it did in the observed data. Samples in the second treatment group were generated by removing 3% of the OTUs from the dataset and recalculating the relative abundance of the remaining OTUs. Sequences were sampled from the richness-adjusted community distrubtion until each sample had the same number of sequences that it did in the observed data. Under the richness-adjusted community model each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups. Because the construction of the richness-adjusted community model was a stochastic process, 100 replicates were geneated for each dataset.

**Test of statistical significance.** Statistical comparisons of alpha diversity metrics across the simulated treatment groups were performed using the non-parametric two-sample Wilcoxon test as implemented in R. This test was selected because the alpha-diversity metrics tended to not be normally distributed and each dataset required a different transformation to normalize the data. Comparisons of beta diversity metrics were performed using the adonis2 function from the vegan (v.`r package_version("vegan")`) R package [@Oksanen2022]. The adonis2 function implements a non-parametric multivariate analysis of variance using distances matrices as described by McArdle and Anderson [@McArdle2001]. Throughout this study I used 0.05 as the threshold for asssessing statistical significance.

**Power analysis.** The parameters used to design the skewed abundance and richness-adjusted community models were set to impose a known effect size when using rarefied community data. The statistical power to detect these differences was determined by calculating the p-value for each of 100 replicate simulated set of samples from each dataset using the various alpha and beta diversity metrics. The percentage of tests that yielded a significant P-value was considered the statistical power (i.e., 1 minus the Type II error) to detect the difference.

@@

The inverse Simpson diversity was calculated as $1/D_{simpson}$. In both formulae, $n_i$ is the number of sequences in OTU $i$ and $N$ is the number of sequences in the sample. Rarefaction of richness, Shannon diversity and Inverse Simpson diversity values were carried out in mothur. Briefly, mothur calculates each value on a random draw of the same number of sequences from each sample and obtains a mean value based on 1,000 random draws. Scaled ranked subsampling (SRS) was used to normalize OTU counts to the size of the smallest sample in each dataset using the SRS R package (v.`r package_version("SRS")`)[@Beule2020]. Normalized OTU counts were used to calculate sample richness and Shannon and inverse Simpson diversity values using mothur. Data normalized by cumulative sum scaling (CSS) were not reported for alpha-diversity values since the relative abundances of the features do not change with the normalization procedure [@Paulson2013]. The non-parametric bias-corrected Chao1 and ACE richness estimators [@Chao2016] and a non-parametric estimator of the Shannon diversity [@Chao2003] were calculated using raw OTU counts with mothur. Parametric estimates of sample richness were calculated using the breakaway (BA) R package (v.`r package_version("breakaway")`)[@Willis2015]. The current analysis reports both the results from running default model selection procedure and the Poisson model. The default model selection returned either the Kemp, Negative Binomial, or Poisson models. Relative abundance data were not used to calculate alpha diversity metrics since the richness and evenness does not change from the raw data when dividing each sample by the total number of sequences in the sample.

@@

These metrics were calculated using the relative abundance of each OTU using the vegdist function from vegan. The relative abundance was calculated as the number of sequences in the OTU (e.g., $n_{A,i}$) divided by the total number of sequences in the sample (e.g., $N_A$).

Rarefied beta-diversity values were calculated using the avgdist function in vegan. Briefly, vegan's avgdist function calculates each pairwise dissimilarity index after obtaining a random draw of the same number of sequences from each sample. After obtaining 100 random draws it returns the mean value.

Three approaches were taken to normalize the number of sequences across samples within a dataset. Scaled ranked subsampling (SRS) and cumulative sum scaling (CSS) were used to normalize raw OTU counts using the SRS (v.`r package_version("SRS")`) and metagenomeSeq (v.`r package_version("metagenomeSeq")`) R packages, respectively [@Beule2020; @Paulson2013]. The normalized counts were then used to calculate Jaccard and Bray-Curtis dissimilarity indices. Finally, the variance-stabilization transformation (VST) as implemented in the DESeq2 (v.`r package_version("DESeq2")`) R package was used to normalize the data as described by McMurdie and Holmes [@Love2014; @McMurdie2014]. Because the VST approach generates negative values, which are incompatible with calculating Jaccard and Bray-Curtis dissimilarity values, Euclidean distances were calcualted instead.

Raw OTU counts were used to calculate centered log ratio values for each OTU, which were then used to calcualte Euclidean distances; such distances are commonly refered to as Aitchison distances. Centered log-ratio (CLR) abundances are calculated as:

$$
clr\left(n_j\right) = \left[ \ln\frac{x_{ij}}{g(x_j)}, ..., \ln\frac{x_{S_Tj}}{g(x_j)}\right]
$$

where $x_{ij} was the number of sequences observed for OTU $i$ in sample $j$ and $g()$ was the geometric mean $x_j$ was the counts of the $S_T$ OTUs in sample $j$. Because the geometric mean is zero if any OTU is absent from a sample, the CLR is undefined when there are unobserved OTUs in a sample. To overcome this problem, I attempted a four approaches. The first, Zero CLR, imputed the value of the zeroes based on the observed data using the zCompositions (v.`r package_version("zCompositions")`) R package [@Palarea2015]. The second, One CLR, added a pseudocount of 1 to the abundance of all OTUs [@Paulson2013; @Lin2020]. The third, Nudge CLR, added a pseudocount of 1 divided by the total number of sequences in a sample to each OTU in the sample [@Costea2014; @Lin2020]. The final approach, Robust CLR, removed unobserved OTUs prior to calcualting the CLR [@Martino2019].

**Analysis of sequencing coverage.** To assess the level of sequencing coverage I calcualted Good's coverage ($C_{Good}$) using mothur:

$$C_{Good} = 100\% \times \left(1-\frac{n_1}{N_T} \right)$$

where $n_1$ is the number of OTUs with only one sequence in the sample and $N_T$ is the total number of sequences in the sample. Good's coverage was calculated using (i) the observed OTU counts for each sample and dataset, (ii) following rarefaction (1,000 iterations) of the observed OTU counts to the size of the smallest sample in each dataset, and (iii) after rarefying (1,000 iterations) the null community distribution.

**Reproducible data analysis.** A complete reproducible workflow written in Snakemake (v.7.15.2) and conda (v.4.12.0) computational environment can be obtained from the GitHub hosted git repository for this project (https://github.com/SchlossLab/Schloss_Rarefaction_XXXXX_2022). This paper was written in R markdown (v.`r package_version("rmarkdown")`) with the kableExtra (v.`r package_version("kableExtra")`) package. The mothur (v.1.47.0) and R (`r glue("{version$major}.{version$minor}")`) software packages were used for all analyses with extensive use of functions in the tidyverse metapackage (v.`r package_version("tidyverse")`). A preliminary version of this analysis was presented as the Rarefaction video series on the Riffomonas Project YouTube channel (https://www.youtube.com/playlist?list=PLmNrK_nkqBpJuhS93PYC-Xr5oqur7IIWf).

\vspace{10mm}

@@

I am grateful to the researchers who generated the datasets used in this study. I also thank the individuals who asked questions and commented on the preliminary version of this project, which was released as a YouTube playlist on the Riffomonas channel. This work was supported in part by funding from the National Institutes of Health (U01AI124255, P30DK034933, R01CA215574).

@@

\newpage
## Tables
